---
title: "Linked-Disturbance-RF"
author: "Xiulin Gao"
email:  "xiulingao@lbl.gov"
date:   "10-May-2023"
---
## Introduction
This script is used to do spatial random forest modeling of above-ground biomass for
the Rocky Mountains using multi-predictors. 

To run this notebook on jupyter notebook on HPC system, do following:
1. convert to .ipynb in terminal: jupytext notebook-name.ipynb --output Linked-Disturbance-RF.Rmd
2. create R environment and install all needed packages and r kernel to run on jupyter notebook
   1) conda create -n my-R-env python=3.9
   2) source activate my-R-env
   3) conda install -c conda-forge mamba # currently the best package manager to solve environment conflicts for R
   4) mamba install -c conda-forge r-raster r-rgeos r-rgal r-tidyverse r-sf r-spatialRF r-randomForestExplainer         r-irkernel  ## any package you want here, just be aware that sometimes the environment cannot be solved. 
   5) launch jupyter notebook and select the R environment you just created to run this notebook.
   


```{r,label="reset-R",message=FALSE,results='hide'}

# Unload all packages except for the R default ones
plist = names(sessionInfo()$otherPkgs)
if (length(plist) > 0){
   dummy = sapply(X=paste0("package:",plist),FUN=detach,character.only=TRUE,unload=TRUE)
}#end if (length(plist) > 0)


# Remove all variables
rm(list=ls())

# Reset warnings
options(warn=0)

# Close all plots
invisible(graphics.off())

# Clean up
invisible(gc())

```




```{r,label='load-packages',message=FALSE,results='hide'}
library(tidyverse)
library(spatialRF)
library(ggplot2)
library(data.table)
library(pcaMethods)
library(sf)


```



```{r,label='path-setting',message=FALSE,results='hide'}
home_path = path.expand("~")
git_path  = file.path(home_path,"linked_disturbance")
data_main = file.path(git_path,"data")
data_file = "linked_disturbance_data_frame.csv"

```




```{r,label='make-dir',message=FALSE,results='hide'}

plot_main    = file.path(git_path,"results")
explor_path  = file.path(plot_main,"explore-plots") 
dummy        = dir.create(explor_path  , recursive = TRUE, showWarnings = FALSE)

```





```{r,label='plot-setting',message=FALSE,results='hide'}

gg_device  = c("png")     # Output devices to use (Check ggsave for acceptable formats)
gg_depth   = 600          # Plot resolution (dpi)
gg_ptsz    = 18           # Font size
gg_ptszl   = 26
gg_width   = 17.5         # Plot width (units below)
gg_widthn  = 14.5   
gg_height  = 8.5          # Plot height (units below)
gg_units   = "in"         # Units for plot size
gg_screen  = TRUE         # Show plots on screen as well?
ndevice    = length(gg_device)

top12 <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
           "#44AA99", "#999933", "#882255", "#6699CC", "#888888", "#661100")




```




```{r,label='predictors-PCA',message=FALSE,results='hide'}

mod_df      = fread(file.path(data_main,data_file))
# select only quantitative columns for RF as spatialRF mainly deals with quantitative data
var_class = split(names(mod_df),sapply(mod_df, function(x) paste(class(x), collapse=" ")))
quant_var = c(var_class$integer,var_class$numeric)
pred_x    = quant_var[-c(1:4,7,17:22,24,47,70,93,116:122)]
pca_df  = mod_df %>% select(all_of(pred_x))
pca_df  = prep(pca_df, scale="uv",center=TRUE)
#varcors = cor(pca_df, method = "pearson", use = "pairwise")
#corrplot::corrplot(varcors)
npc     = 19
fppca   = pca_df %>% pca(nPcs=npc, method="nipals") 
summary(fppca)
#biplot(fppca)
fppca.loads = as.data.frame(loadings(fppca))

### given PCA loading select some independent variables
xs    = rownames(fppca.loads)
pc_x  = list()
for (p in sequence(npc)){
  x_p  = xs[fppca.loads[p] > 0.2 | fppca.loads[p] < -0.2]
  pc_x = append(pc_x,x_p)
}

pc_x  = unlist(pc_x)
x_var = pc_x[!duplicated(pc_x)]

rf_df = mod_df %>% select(all_of(c(x_var,"agbd","lat_lowest","lon_lowest","utm_z13n_easting","utm_z13n_northing")))


```




```{r,label='model-data-check',message=FALSE,results='hide'}
## the spatialRF package does have special requirements for model data, so check first

if (sum(apply(rf_df, 2, is.na)) > 0){rf_df = na.omit(rf_df)} # remove rows withn NAs
zerosd_col  = names(apply(rf_df, 2, var) == 0)[apply(rf_df, 2, var) == 0] 
na_find     = sapply(as.data.frame(scale(rf_df)), function(x)any(is.nan(x)))
na_col      = names(na_find)[na_find] 
inf_find    = sapply(as.data.frame(scale(rf_df)), function(x)any(is.infinite(x)))
inf_col     = names(inf_find)[inf_find] 

cat(paste0("columns with zero variance are",zerosd_col))
cat(paste0("columns produce NAs when scale are",na_col))
cat(paste0("columns produce Inf when scale are",inf_col))

```





```{r,label='multicollinearity-and-interactions',message=FALSE,results='hide'}

y_var = "agbd"
x_var = x_var #currently not include lon and lat as predictors
cords = rf_df[, c("utm_z13n_easting","utm_z13n_northing")]
names(cords)   = c("x","y")
dist_matrix    = as.matrix(dist(cords, diag=TRUE, upper=TRUE))
dist_threshold = c(0, 20000, 40000, 80000, 160000) # in same unit as the dist.matrix, which assume is m 
seed           = 1

# remove predictors that are highly correlated
preference.order         = c( "elevation"
                            , "speiCum5yrPrior"
                            , "modisTreeVeg"
                            , "chili"
                            , "aspect"
                            , "slope"
                            )
x_df                     = rf_df %>% select(all_of(x_var))
predictor.variable.names =  auto_cor( x = x_df,
                                      cor.threshold = 0.6,
                                      preference.order = preference.order
                                    ) %>% 
                            auto_vif( vif.threshold = 2.5,
                                      preference.order = preference.order
                                    )

x_update    = predictor.variable.names$selected.variables
rfdf_update = rf_df %>% select(all_of(c("utm_z13n_easting","utm_z13n_northing",x_update,"agbd")))

## interactions, this can run for long and out of memory space
## suggest running this on HPC, NEVER run it on local machine 

local.cluster  = parallel::makeCluster(parallel::detectCores() - 4,
                                       type = "PSOCK")
doParallel::registerDoParallel(cl = local.cluster)

interactions = the_feature_engineer( data                     = rfdf_update
                                   , dependent.variable.name  = y_var
                                   , predictor.variable.names = x_update
                                   , xy                       = cords
                                   , importance.threshold     = 0.50 
                                   , cor.threshold            = 0.60 
                                   , seed                     = seed
                                   , repetitions              = 100
                                   , verbose                  = TRUE
                                   , cluster                  = local.cluster
                                   )

parallel::stopCluster(cl = local.cluster)

rfdf_update = interactions$data
x_update    = interactions$predictor.variable.names

```





```{r,label='random-forest',message=FALSE,results='hide'}

local.cluster  = parallel::makeCluster(parallel::detectCores() - 4,
                                       type = "PSOCK")
doParallel::registerDoParallel(cl = local.cluster)

# non-spatial model
agb_mod_nonsp  = rf( data                     = rfdf_update
                   , dependent.variable.name  = y_var
                   , predictor.variable.names = x_update
                   , distance.matrix          = dist_matrix
                   , distance.thresholds      = dist_threshold
                   , xy                       = coords
                   , seed                     = seed
                   , verbose                  = FALSE
                   , cluster                  = local.cluster
                   )

plot_residuals_diagnostics( agb_mod_nonsp
                          , verbose = FALSE)
plot_importance( agb_mod_nonsp
                ,verbose = FALSE)

# spatial model
agb_mode_sp  =  rf_spatial( model   = agb_mod_nonsp
                          , method  = "mem.moran.sequential"
                          , verbose = FALSE
                          , seed    = seed
                          , cluster = local.cluster
                          )                           %>% 
                rf_tuning()                           %>% 
                rf_evaluate()                         %>% 
                rf_repeat()

parallel::stopCluster(cl = local.cluster)

# importance comparison

p1 =  plot_importance( agb_mod_nonsp 
                     , verbose = FALSE) + 
      ggplot2::ggtitle("Non-spatial model") 

p2 =  plot_importance( agb_mod_sp
                     , verbose = FALSE) + 
      ggplot2::ggtitle("Spatial model")

p1 | p2 



```

